Foregrounds in Fourier Space for Redshifted 21~cm Observatories

### 2015-02-05 ms.tex
Start rewriting section 5
from 48d0c33:
============

Add figure6a.eps, figure6b.eps to git version control

In ms.tex,

(*) Start rewriting section 5

\section{Example Data}\label{sec:data}

@@

MWA construction was completed in 2012 and, after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles, each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg~ wide and multiple secondary lobes. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\boldsymbol{b}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz. The center frequency of 185~MHz corresponds to a redshift, $z\simeq 6.68$. {\bf At this frequency, we use $A_\textrm{e}=10.5$~m$^2$}.

@@

The MWA targets two primary low-foreground fields for reionization observations. Here, we focus on the field at RA = 0$^\textrm{h}$, Dec = $-30\arcdeg$. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of the dipoles in an MWA tile. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg~ grid. During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary (once in $\sim 30$~minutes). This process is repeated throughout the course of the observation $\approx 4.86$~hours. Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

The observations shown here were undertaken on 2013~August~23. % Data set from this day has been chosen as a reference for comparison within the MWA EoR collaboration because this is one of the days featuring a low number of known system errors from among a long stretch of stable observing.  
We have chosen two, 112 second long sections from this night for detailed study. These were chosen to provide a selection of possible foreground and instrumental conditions. As an example of a nominal observing setup we choose a zenith pointing; as an example of poor foreground conditions, we choose a pointing when the field is $\sim 2$~hours from zenith. This pointing has a significantly higher secondary lobe structure and is observed when the bright galactic center is well above the horizon.

The {\it local sidereal time} (LST) at these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at {\bf 22.08~hours (wrapped to $-1.92$~hours in the figure)} and 0.09~hours, which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings, respectively.

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figures/v1_0/custom_pointings.eps}
% \includegraphics[width=\linewidth]{f2.eps}
\includegraphics[width=\linewidth]{fig3.eps}
\caption{MWA tile beam pointing directions during the course of the observation. The $x$--axis refers to the {\it local sidereal time} (LST). The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA). Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg~ or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted lines respectively. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 0$^\textrm{h}$, Dec = $-30$\arcdeg. Dot--dashed vertical lines show the LST of {\it off-zenith} and {\it zenith} pointings used in our study. \label{fig:pointings}}
\end{figure}

@@

\section{Simulations}\label{sec:modeling}

We describe the instrumental and foreground models we use in our simulations.

The instrument model consists primarily of the tile power pattern, $A(\hat{\boldsymbol{s}},f)$ (see equation~\ref{eqn:obsvis}). It is modeled as a mutually--coupled 4--by--4 dipole array with the overall power pattern of each individual dipole calculated via finite element electromagnetic simulation \citep{sut14_arXiv}. To speed up simulations, we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the full simulation, hence we use the idealized dipoles. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known repeatability and stability level of the analog signal chain \citep{bow07b}. Besides having the effect of adding a time--dependent uncertainty in the power pattern, these random delay fluctuations reduce the coherence in the phased addition of dipole signals resulting in deviations from predicted models of the power pattern, most prominently at its nulls. The power patterns so obtained are shown in Figure~\ref{fig:power_pattern} for the pointings chosen in our study.

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figures/v1_0/delta_array_powerpattern_0.3m_ground_custom.eps}
% \includegraphics[width=\linewidth]{f4.eps}
\includegraphics[width=\linewidth]{fig5.eps}
\caption{Models of MWA tile power pattern for {\it off--zenith} (top) and {\it zenith} pointings (bottom) at 185~MHz. An MWA tile is modeled as a 4$\times$4 array of isotropic radiators placed 0.3~m above an infinite ground plane. Random fluctuations of rms 0.05~ns have been added to delay corrections during the phased addition of voltages from these isotropic radiators. The logarithmic color scale (shown on the right) is common to both panels. \label{fig:power_pattern}}
\end{figure}

@@

The MWA has a very wide field of view ($\gtrsim 20\arcdeg$~ at 185~MHz) and has non--negligible response all the way to the horizon. As already noted, foreground objects fill the {\it foreground wedge} with an amplitude which depends on the tile power gain in that direction, with objects at the horizon being closest to the boundaries of the {\it wedge}.

The degree to which these objects leak power into the spectral line--of--sight modes in the {\it EoR window} will be shown to depend primarily on the brightness of foregrounds near the boundaries of the {\it foreground wedge}. Thus, it is important to consider an all--sky model for foreground objects in evaluating the features seen in the power spectrum instead of restricting only to the primary field of view, a point also supported by Pober~et~al.~(2015, in preparation).

We use an all--sky foreground emission model that includes both diffuse and bright compact components.

For the diffuse component, we use an all--sky radio foreground model \citep{deo08} to estimate the emission at 185~MHz. At this frequency, since this map is predominantly based on the 408~MHz map of \citet{has82} which has an angular resolution of 0.85\arcdeg, we smoothed the 185~MHz map to the same resolution. However, to avoid any artifacts from sampling this map, we sample it at $\approx 27$\arcmin~intervals. We model the diffuse foreground spectra with a single spectral index at each pixel in the map, found by comparing model maps at 170~MHz and 200~MHz.

A low resolution version of the diffuse foreground model used is shown in Figure~\ref{fig:DSM}. Contours of the MWA tile power pattern shown in Figure~\ref{fig:power_pattern} are also overlaid. Of notable significance in the {\it off--zenith} pointing is the presence of a portion of the Galactic plane and the bright Galactic center in the westward sky, where the MWA tile power gain is significant ($\gtrsim 12$\%). In the {\it zenith} pointing, the Galactic plane has set and the power pattern in that direction is at least 16 times smaller.

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figures/v1_0/dsm.eps}
% \includegraphics[width=\linewidth]{f5.eps}
\includegraphics[width=\linewidth]{fig6.eps}
\caption{Sky brightness temperature of the diffuse foreground model at 185~MHz visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels. Contours of power pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.00195, 0.00781, 0.0312, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing and the MWA tile power gain is significant ($\gtrsim 12$\%) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly lesser. \label{fig:DSM}}
\end{figure}

The model described in the preceding section is primarily a model of the diffuse foreground sky. While it contains faint compact emission blended in with the diffuse emission, {\bf bright point sources have been removed \citep{deo08}}. In order to supplement it with missing bright compact emission, we use classical radio source confusion estimates to determine the nominal flux density threshold and include point sources brighter than this threshold. It may be noted that slightly different criteria are in common use in radio astronomy to estimate radio source confusion \citep[see Appendix of][and references therein]{thy13}. For an angular resolution of 0.85\arcdeg, using a conservative `$S_\textrm{c}=5\sigma_\textrm{c}$' criterion, we determine the flux density threshold to be $\approx 10$~Jy. Other liberal criteria that yield a lower threshold carry a greater risk of double--counting point sources which might be already blended in with the diffuse sky model.

From the SUMSS catalog, we select {\bf objects} whose deconvolved major axes are equal to 0\arcsec, {\bf thereby strictly selecting point sources}. From the NVSS catalog, we excluded objects that overlap with those in the SUMSS survey footprint. Point sources from NVSS were selected if the convolved major axes were not greater than $\approx 47$\arcsec, which matches the angular resolution of the survey. Using a mean spectral index of $\langle\alpha_\textrm{sp}\rangle=-0.83$ (flux density, $S(f)\propto f^{\alpha_\textrm{sp}}$) obtained by \citet{mau03} for both NVSS and SUMSS catalog objects, we calculate the corresponding flux densities at 185~MHz, $S_{185}$. From this subset, we choose point sources with $S_{185}\geq 10$~Jy. The selection of such bright point sources is not affected by minor differences in flux density sensitivity of the two surveys. We verified that our selection criteria ensure a similar areal density of objects in the two surveys.

Based on these criteria, we selected 100 objects from the SUMSS catalog and 250 objects from the NVSS catalog. Together with the diffuse foreground model, we obtain an all--sky foreground model consisting of both compact and diffuse emission. Figure~\ref{fig:CSM} shows the compact foreground emission model used in our study whose with contours of the power pattern overlaid. The flux densities of the point sources are indicated by the color scale.

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figures/v1_0/csm.eps}
% \includegraphics[width=\linewidth]{f6.eps}
\includegraphics[width=\linewidth]{fig7.eps}
\caption{Flux densities of bright point sources at 185~MHz visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels and corresponds to the color--coded filled circles. Contours of power pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.001953125, 0.0078125, 0.03125, 0.125, and 0.5. \label{fig:CSM}}
\end{figure}

It is worth emphasizing that the ideal foreground model is one with an angular resolution finer than that of the instrument that covers the whole sky at the observing frequency, a point we will revisit later. For this study, we note that our conclusions are not significantly affected despite minor differences in the thresholds from different radio source confusion estimates.

>>>>

\section{MWA Data Modeling}\label{sec:data}

We now use our simulations to analyze features in the delay power spectrum in data obtained using the MWA.

@@

MWA construction was completed in 2012 and, after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles, each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg~ wide and multiple secondary lobes. The array layout is described in \citet{bea12}. %{\bf At the observing frequency of 185~MHz, we use $A_\textrm{e}=10.5$~m$^2$}.

@@

The MWA targets two primary low-foreground fields for reionization observations. Here, we focus on the field at RA = 0$^\textrm{h}$, Dec = $-30\arcdeg$. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of the dipoles in an MWA tile. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg~ grid. During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary (once in $\sim 30$~minutes). This process is repeated throughout the course of the observation $\approx 4.86$~hours. % Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

The observations {\bf used} here were undertaken on 2013~August~23. We have chosen two, 112 second long sections from this night for detailed study. These were chosen to provide a selection of possible foreground and instrumental conditions. As an example of a nominal observing setup we choose a zenith pointing; as an example of poor foreground conditions, we choose a pointing when the field is $\sim 2$~hours from zenith. This pointing has a significantly higher secondary lobe structure and is observed when the bright galactic center is well above the horizon.

These two pointings are at LST 22.08~hours and 0.09~hours, which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings, respectively.

% The {\it local sidereal time} (LST) at these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at {\bf 22.08~hours (wrapped to $-1.92$~hours in the figure)} and 0.09~hours, which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings, respectively.

% \begin{figure}[htb]
% \centering
% % \includegraphics[width=\linewidth]{figures/v1_0/custom_pointings.eps}
% % \includegraphics[width=\linewidth]{f2.eps}
% \includegraphics[width=\linewidth]{fig3.eps}
% \caption{MWA tile beam pointing directions during the course of the observation. The $x$--axis refers to the {\it local sidereal time} (LST). The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA). Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg~ or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted lines respectively. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 0$^\textrm{h}$, Dec = $-30$\arcdeg. Dot--dashed vertical lines show the LST of {\it off-zenith} and {\it zenith} pointings used in our study. \label{fig:pointings}}
% \end{figure}

@@

\subsection{Modeling}\label{sec:modeling}

The MWA tile power pattern is modeled as a mutually--coupled 4--by--4 dipole array with the overall power pattern of each individual dipole calculated via finite element electromagnetic simulation \citep{sut14_arXiv}. To speed up simulations, we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the full simulation, hence we use the idealized dipoles. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known repeatability and stability level of the analog signal chain \citep{bow07b}. Besides having the effect of adding a time--dependent uncertainty in the power pattern, these random delay fluctuations reduce the coherence in the phased addition of dipole signals resulting in deviations from predicted models of the power pattern, most prominently at its nulls. % The power patterns so obtained are shown in Figure~\ref{fig:power_pattern} for the pointings chosen in our study.

We use the model described in \S\ref{sec:foreground} for the foreground sky. Figure~\ref{fig:sky-model-MWA} shows the diffuse emission and bright point source foreground models for the two chosen pointings with the modeled MWA tile power pattern contours overlaid. Notice the presence of a portion of the Galactic plane and the bright Galactic center in the westward sky in the diffuse sky model, where the MWA tile power gain is significant ($\gtrsim 12$\%). In the {\it zenith} pointing, the Galactic plane has set and the power pattern in that direction is at least 16 times smaller.

\begin{figure*}[htb]
\centering
\subfloat[][Diffuse Sky Model]{\label{fig:DSM}\includegraphics[width=0.45\linewidth]{figure6a.eps}}
\subfloat[][Bright Point Source Model]{\label{fig:CSM}\includegraphics[width=0.45\linewidth]{figure6b.eps}}
\caption{Sky brightness temperature of the diffuse foreground model (left) and flux densities of bright point sources (right) at 185~MHz visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scales are logarithmic. Power pattern contours are overlaid. The contour levels shown are 0.00195, 0.00781, 0.0312, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing in the diffuse sky model and the MWA tile power gain is significant ($\gtrsim 12$\%) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly lesser.}
\label{fig:sky-model-MWA}
\end{figure*}

@@

% It is worth emphasizing that the ideal foreground model is one with an angular resolution finer than that of the instrument that covers the whole sky at the observing frequency, a point we will revisit later. For this study, we note that our conclusions are not significantly affected despite minor differences in the thresholds from different radio source confusion estimates.
